Union and intersection contracts are hard, actually

### 2021-06-09 paper.tex
Pass on the union contract subsection

@@

\lstset{
  escapeinside={(*@}{@*)} % escape sequence to insert latex
}
\newcommand{\nickel}[1]{\lstinline[language=nickel]{#1}}
\newcommand{\racket}[1]{\lstinline[language=racket]{#1}}
\newcommand{\haskell}[1]{\lstinline[language=Haskell]{#1}}
\newcommand{\typescript}[1]{\lstinline[language=JavaScript]{#1}}

@@

Union types, meaning a type \nickel{A @| B} containing values which
belong either to a type \nickel{A} or \nickel{B}, are a popular tool
when adding a static types to a dynamic language. In particular, both
TypeScript~\cite{TypeScriptUnions} and MyPy~\cite{MyPyOptional}, use
union types to model the frequent practice to use the value
\lstinline{null} (\lstinline{None} in Python) to represent an absent
optional value. This is why the gradual typing literature, concerned
with formalising the interplay between static and dynamic type
systems, has been quite interested in union
types~\cite{RootCauseOfBlame,KeilThiemannUnionIntersection}.

@@

\subsection{Unions}
Unions ($\cup$) are a fundamental building block of programs.
A union holds a value whose
shape is known to be in a fixed set of alternatives, without exposing
which of those alternatives the value belongs to.
For example, a value that could be either a Number or a String can be
modelled using unions as \nickel{Num @| String}.
They correspond logically to disjunctions.

Statically typed languages, and more specifically functional programming
languages, provide a similar abstraction, sums
(also called tagged unions, or discriminated unions), which
also allow to model a value whose shape conforms to a fixed set of
alternatives, but requiring an explicit mark (or tag) that differentiates
to which of these alternatives it belongs to.
For example, consider the Haskell code presented on Figure
\ref{fig:safediv-haskell}, it's a safe division that will
check if the dividend is 0, and if it is, return an
error message.
Since the function can return two different types, a sum
(\haskell{Either}) is used, and on each
of the if branches an explicit tag has to be used (either
\haskell{Left} or \haskell{Right}).

\begin{figure}[h]
\begin{lstlisting}[language=Haskell]
safeDiv :: Int -> Int -> Either String Int
safeDiv x y =
  if y == 0 then
    Left "Division by 0 is forbidden"
  else
    Right (div x y)
\end{lstlisting}
\caption{Safe division implemented in Haskell}
\label{fig:safediv-haskell}
\end{figure}

This incurred verbosity certainly facilitates the work of static checks,
like typechecking, but dynamically typed languages permit much more flexibility,
and programmers don't usually expect to tag their values.
That is why, as mentioned on the Introduction, when designing
type systems for dynamically typed languages, unions become of particular
interest.
See the implementation of a safe division function on TypeScript
(Figure \ref{fig:safediv-ts}).
Neither of the if branches has to tag their return value, and the return
type of the function is modelled using unions, \typescript{string | number}.

\begin{figure}[h]
\begin{lstlisting}[language=JavaScript]
function safeDiv(x: number, y: number): string | number {
  if (y == 0) {
      return "Division by 0 is forbidden"
  } else {
      return x/y
  }
}
\end{lstlisting}
\caption{Safe division implemented in TypeScript}
\label{fig:safediv-ts}
\end{figure}

% \begin{figure}[h]
% \begin{lstlisting}[title={Tree-like data structure}]
% type BinaryTree =
% Integer @| (BinaryTree, BinaryTree)
% \end{lstlisting}
% \caption{A binary tree data type using unions}
% \label{fig:union-adt-ex}
% \end{figure}

Unions are not only useful to model the outcome of alternative
branches of computations, but they can be used to
formalize very common patterns of data types:
\begin{itemize}
    \item To represent nullable, or optional, values.
\begin{lstlisting}[language=nickel]
optionalValue : Str @| Null
\end{lstlisting}

    \item For plain enumerations of literals.
\begin{lstlisting}[language=nickel]
protocol : "Http" @| "Ftp" @| "Sftp"
\end{lstlisting}

    \item To support different representations of the same data.
\begin{lstlisting}[language=nickel]
person : {name : Str, age : Int}
  @| {name : Str, dateOfBirth : Date}
\end{lstlisting}

\end{itemize}
All of which are common in configurations.

% \begin{figure}[h]
% % \label{fig:union-nullable-ex}
% \begin{lstlisting}[title={Nullable values}]
% optionalValue : Str @| Null
% \end{lstlisting}

% % \label{fig:union-enum-ex}
% \begin{lstlisting}[title={Plain enumeration}]
% protocol : `Http | `Ftp | `Sftp
% \end{lstlisting}

% % \label{fig:union-alt-ex}
% \begin{lstlisting}[title={Alternative representations}]
% person : {name : Str, age : Int}
%   @| {name : Str, dateOfBirth : Str}
% \end{lstlisting}
% \caption{Various use cases of unions}
% \label{fig:unions-use-cases}
% \end{figure}

% In statically typed languages, and more specifically in functional programming
% languages, sums (also called tagged unions, or discriminated unions) are usually
% preferred, as constructors for algebraic datatypes. [references]. Sums are unions' cousin where the alternatives are
% syntactically separated by an associated discriminating label. Well-known
% programming languages with first-class tagged unions include Haskell, OCaml,
% Rust, Scala, PureScript, Elm, Swift and many more.

% On the other hand, in dynamically typed languages, it is customary to directly
% store different kinds of data in one variable without further ceremony. The
% runtime system needs to attach type tags to values anyway, which can usually be
% observed using functions like JavaScript's \verb+typeof+. Using such bare
% unions\resolved{If you can observe with typeof, then it's pretty much a tagged
% union, though the tag is implicit} has already been an idiom in dynamically
% typed language for a long time. Being able to type and handle these unions is an
% important aspect of the migration to a gradual type system. Indeed, a lot of
% \resolved{This is the first time discussing gradual type systems, maybe it should
% be introduced earlier, in particular how they relate to contracts
% (I think is mentioned on the appendix)}
% gradually typed languages either existed originally as dynamically typed
% language or aim at backward compatibility with a dynamically typed base
% language. This is why despite the presence of a static type system, gradually
% typed languages tend to favor bare unions types which is the typed counterpart
% of this practice (as e.g in TypeScript \cite{TypeScriptUnions}).

Nickel has no pre-existing untyped language that we have to
support, so we are freer of our choice.
However, Nickel is a configuration language, whose contract and type system
is intended to encode data schemas encountered in the wild. Since these schemas do not use
tags to differentiate different kinds of values, there is a strong incentive to support untagged unions at the
type level, or at least, at the contract level.

We have shown how values of a union are built, is worth mentioning
how they are used.
For instance, consider the TypeScript function on Figure \ref{fig:addElem},
we want it to insert an element to a list that is either an
\typescript{Array<string>} or an \typescript{Array<number>},
what value could we fill \typescript{...} with for this to typecheck?
Since we don't know what's the shape of \typescript{list},
we can't know whether to push a \typescript{string} or a
\typescript{number}.
That is, when using a union, we have to treat it as if it where any one
of the alternatives, in the case of the example, we should provide
it with a value that is both a \typescript{string} and a \typescript{number}.

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let Animal = { species: Str, race: Str } in
let Pet = { owner: Str, name: Str } in
let myDog | Animal & Pet =
  { species = "Canis Lupus",
    race = "Australian Cattle Dog",
    owner = "Anonymous Author",
    name = "Juno" }
\end{lstlisting}
\caption{An animal that is also a pet}
\label{fig:intersection-record}
\end{figure}

@@

% LocalWords:  Felleisen JSON TOML YAML templating Jsonnet Dhall MyPy
% LocalWords:  Kubernetes DevOps natively nixpkgs disjunctions
% LocalWords:  TypeScript

>>>>

\lstdefinelanguage{Python}{
  % keywords={typeof, new, true, false, catch, function, return, null, catch, switch, var, if, in, while, do, else, case, break},
  % keywordstyle=\color{blue}\bfseries,
  % ndkeywords={class, export, boolean, throw, implements, import, this},
  % ndkeywordstyle=\color{darkgray}\bfseries,
  identifierstyle=\color{black},
  sensitive=false,
  % comment=[l]{//},
  % morecomment=[s]{/*}{*/},
  % commentstyle=\color{purple}\ttfamily,
  % stringstyle=\color{red}\ttfamily,
  morestring=[b]',
  morestring=[b]"
}

@@

\lstset{
  escapeinside={(*@}{@*)} % escape sequence to insert latex
}
\newcommand{\nickel}[1]{\lstinline[language=nickel]{#1}}
\newcommand{\racket}[1]{\lstinline[language=racket]{#1}}
\newcommand{\haskell}[1]{\lstinline[language=Haskell]{#1}}
\newcommand{\typescript}[1]{\lstinline[language=JavaScript]{#1}}
\newcommand{\python}[1]{\lstinline[language=Python]{#1}}

@@

Union types, meaning a type \nickel{A @| B} containing values which
belong either to a type \nickel{A} or \nickel{B}, are a popular tool
when adding a static types to a dynamic language. In particular, both
TypeScript~\cite{TypeScriptUnions} and MyPy~\cite{MyPyOptional}, use
union types to model the frequent practice to use the value
\typescript{null} (\python{None} in Python) to represent an absent
optional value. This is why the gradual typing literature, concerned
with formalising the interplay between static and dynamic type
systems, has been quite interested in union
types~\cite{RootCauseOfBlame,KeilThiemannUnionIntersection}.

@@

\subsection{Unions}

A union type \nickel{A @| B} is a type of values which are either of
type \nickel{A} or of type \nickel{B}: literally the union of
\nickel{A} and \nickel{B}. Union types are popular in gradual typed
systems such as TypeScript~\cite{TypeScriptUnions} and
MyPy~\cite{MyPyOptional}.

\paragraph{In gradually typed systems}

In these practical gradual type systems the problem is to be able to
capture, in static types, as many programming patterns as possible
from the underlying dynamically typed language. One such pattern is
heterogeneous collections, in JavaScript (for TypeScript) and Python
(for MyPy), heterogeneous collections, say arrays with values of
different types, are frequent. For instance, in TypeScript, an array
which can contain both strings and numbers would have type
\typescript{Array<string|number>}.

A probably even more common pattern is a variable which can contain
either a value of type a (say, a number) or the \typescript{null}
value. So much so, in fact, that MyPy defines a type alias
\python{Optional[A]} for \python{Union[A,None]} (\python{None} is how
Python renders the null value).

Yet another application of union types is, rather than capturing a
pattern from JavaScript or Python, to capture a pattern from
traditional statically typed language: sum types. In statically typed
languages, values of sum types are usually thought of as being built
out of constructors. But neither JavaScript nor Python have such
constructors. So instead, sums are construed as ``tagged unions'' (or
discriminated unions), that is, quite literally, the union of two
types which contain a discriminating tag. See
Figure~\ref{fig:tagged-union} for an example from the TypeScript
documentation: there the \typescript{kind} field is the tag, and its
type in both alternatives is a singleton type which contains only the
specified string.

\begin{figure}
  \centering
  \begin{lstlisting}[language=JavaScript]
    interface Circle { kind: "circle"; radius: number; }

    interface Square { kind: "square"; sideLength: number; }

    type Shape = Circle | Square;
  \end{lstlisting}

  \caption{A sum type as a tagged union}
  \label{fig:tagged-union}
\end{figure}

\paragraph{Union contracts}

In the academic gradual type literature, it is common to use contracts
as a glue between static and dynamic types. Therefore, the question of
bringing union to contracts is natural, and have indeed been studied
(\emph{e.g.}~\cite{KeilThiemannUnionIntersection,RootCauseOfBlame}).

Like for static types, a valid value for the contract \nickel{A @| B}
is a value which is valid either for contract \nickel{A} or for
contract \nickel{B} (though in Section~\ref{sec:issues-literature} we
will see that it may be desirable to weaken this definition).

Nickel is a language built from scratch with contracts, so it may be
less clear why unions are useful. However, Nickel's ambition is to
have its data model canonically interpretable in common serialization
formats, in particular JSON. It means that it is very convenient to
represent optional value by the \nickel{null} value like in
JavaScript. It also means that Nickel doesn't have built-in
constructors: constructors don't have a canonical representation in
JSON. So it would be quite natural to represent optional contracts and
sum contracts as unions.

% Arnaud: I'm not deleting this quite yet because we may want to use
% these examples, I just didn't manage to make them fit in the flow of
% the narrative.
%
% Unions are not only useful to model the outcome of alternative
% branches of computations, but they can be used to
% formalize very common patterns of data types:
% \begin{itemize}
%     \item To represent nullable, or optional, values.
% \begin{lstlisting}[language=nickel]
% optionalValue | Str @| Null
% \end{lstlisting}

%     \item For plain enumerations of literals.
% \begin{lstlisting}[language=nickel]
% protocol | "Http" @| "Ftp" @| "Sftp"
% \end{lstlisting}

%     \item To support different representations of the same data.
% \begin{lstlisting}[language=nickel]
% person | {name : Str, age : Int}
%   @| {name : Str, dateOfBirth : Date}
% \end{lstlisting}

% \end{itemize}
% All of which are common in configurations.

\unsure{Arnaud: this paragraph is pretty good, on the other hand, there
is no value which lets me complete the example, so this example falls
a bit flat.}
We have shown how values of a union are built, is worth mentioning
how they are used.
For instance, consider the TypeScript function on Figure \ref{fig:addElem},
we want it to insert an element to a list that is either an
\typescript{Array<string>} or an \typescript{Array<number>},
what value could we fill \typescript{...} with for this to typecheck?
Since we don't know what's the shape of \typescript{list},
we can't know whether to push a \typescript{string} or a
\typescript{number}.
That is, when using a union, we have to treat it as if it where any one
of the alternatives, in the case of the example, we should provide
it with a value that is both a \typescript{string} and a \typescript{number}.

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let Animal = { species: Str, race: Str } in
let Pet = { owner: Str, name: Str } in
let myDog | Animal @& Pet =
  { species = "Canis Lupus",
    race = "Australian Cattle Dog",
    owner = "Anonymous Author",
    name = "Juno" }
\end{lstlisting}
\caption{An animal that is also a pet}
\label{fig:intersection-record}
\end{figure}

@@

% LocalWords:  Felleisen JSON TOML YAML templating Jsonnet Dhall MyPy
% LocalWords:  Kubernetes DevOps natively nixpkgs disjunctions
% LocalWords:  TypeScript serializable serialization
